---
title: "Newborn survival"
description: |
  Determining the survival of newborns after birth is essential for not only pediatricians but also obstetricians. This post outlines the survival probability of newborns born and admitted to the Mother and Babies Unit of a tertiary hospital in Ghana.   
author:
  - name: Samuel Blay Nguah
    url: https://sbnguah.netlify.app/
date: 2023-09-22
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
rm(list = ls())
```

### Background 
Predicting the survival of newborns is crucial in clinical decision making, especially prior to deliveries. For instance, the probability of survival of a child born at any gestational age, weight, and a specific locality should be readily known to facilitate clinical judgments and counselling prospective parents.  

### Objective
This post analyses and outlines the probability of survival until discharge of newborns admitted to a tertiary hospital in Ghana. 

### Data
Data for all newborns admitted to the hospital's Mother and Baby Unit was recorded from the year 2012 through 2015. The probability of survival with the 95% confidence interval is as displayed below for males and female separately.

```{r}
df_newborns <-
    readxl::read_xlsx("C:\\Dataset\\mbu2.xlsx") %>% 
    select(wt, age, sex, place, outcome, gestage) %>% 
    filter(place == "KATH" & age < 3) %>% 
    drop_na(gestage) %>% 
    select(-place, -age) %>% 
    filter(
        gestage >=25 & gestage < 49 & wt > 0.6 & wt <=5 & outcome != "Missing" & 
            sex != "Missing") %>%
    mutate(
        sex = factor(sex, levels = c("Female", "Male")),
        outcome = factor(outcome, levels = c("Died", "Discharged")))
```


```{r}
glm1 <- 
    df_newborns %>% 
    glm(outcome ~ wt*sex, family = "binomial", data = .)
```

#### Survival by birth weight and sex 

```{r}
xx <- 
    glm1 %>% 
    ggeffects::ggpredict(terms = c("wt[all]", "sex"))

# data.frame(
#     weight = xx$x, 
#     prob = round(xx$predicted*100, 1), 
#     conf_low = round(xx$conf.low*100, 1), 
#     conf_high = round(xx$conf.high*100,1), 
#     sex = xx$group) %>% 
#     pivot_wider(
#         id_cols = weight, 
#         names_from = sex, 
#         values_from = c(prob, conf_low, conf_high)) %>%
#     mutate(
#         `Birth Weight (kgs)` = sprintf(fmt = "%.1f", weight),
#         Female = paste0(
#             sprintf(fmt = "%.1f", prob_Female), " [", 
#             sprintf(fmt = "%.1f", conf_low_Female), " - ", 
#             sprintf(fmt = "%.1f", conf_high_Female), "]"),
#         Male = paste0(
#             sprintf(fmt = "%.1f", prob_Male), " [", 
#             sprintf(fmt = "%.1f", conf_low_Male), " - ", 
#             sprintf(fmt = "%.1f", conf_high_Male), "]")) %>% 
#     select(`Birth Weight (kgs)`, Female, Male) %>% 
#     as_tibble() %>% 
#     DT::datatable(
#         rownames = FALSE)
```



```{r}
glm1 %>% 
    ggeffects::ggpredict(terms = c("wt[all]", "sex")) %>% 
    plot() +
    labs(title = "Percentage survival by sex",
         x = "Birth weight in kilograms",
         y = "Probability of survival",
         color = "Sex")+
    theme_bw()
    
```

#### Survival by gestational age and sex

```{r}
glm1 <- 
    df_newborns %>% 
    glm(outcome ~ gestage*sex, family = "binomial", data = .)
```

```{r}
xx <- 
    glm1 %>% 
    ggeffects::ggpredict(terms = c("gestage[all]", "sex"))

# data.frame(
#     gestage = xx$x, 
#     prob = round(xx$predicted*100, 1), 
#     conf_low = round(xx$conf.low*100, 1), 
#     conf_high = round(xx$conf.high*100,1), 
#     sex = xx$group) %>% 
#     pivot_wider(
#         id_cols = gestage, 
#         names_from = sex, 
#         values_from = c(prob, conf_low, conf_high)) %>%
#     mutate(
#         `Gestational Age (weeks)` = sprintf(fmt = "%.1f", gestage),
#         Female = paste0(
#             sprintf(fmt = "%.1f", prob_Female), " [", 
#             sprintf(fmt = "%.1f", conf_low_Female), " - ", 
#             sprintf(fmt = "%.1f", conf_high_Female), "]"),
#         Male = paste0(
#             sprintf(fmt = "%.1f", prob_Male), " [", 
#             sprintf(fmt = "%.1f", conf_low_Male), " - ", 
#             sprintf(fmt = "%.1f", conf_high_Male), "]")) %>% 
#     select(`Gestational Age (weeks)`, Female, Male) %>% 
#     as_tibble() %>% 
#     DT::datatable(
#         rownames = FALSE)
```


This is also illustrated graphically below

```{r}
glm1 %>% 
    ggeffects::ggpredict(terms = c("gestage[all]", "sex")) %>% 
    plot() +
    labs(title = "Percentage survival by gestational age in weeks",
         x = "Gestational Age (weeks)",
         y = "Probability of survival",
         color = "Sex")+
    theme_bw()
    
```


#### Survival by gestational age and birth weight simultaneously

Since both gestational age and birth weight have significant effect, we combine 
the two and generate a composite table of percentage survival for males first 
and then females.

```{r}
df_males <- df_newborns %>% filter(sex == "Male")
df_females <- df_newborns %>% filter(sex == "Female")
```


```{r}
glm_3 <- 
    glm(outcome ~ wt*gestage, data = df_males, family = binomial)
```

```{r}
temp <- 
    glm_3 %>% 
    ggeffects::ggpredict(terms = c("wt[all]", "gestage[all]")) 
```

```{r}
tbl_1 <- 
    data.frame(
    wgt = temp[[1]],
    survival = round(temp[[2]]*100,1),
    gest_age = temp[[6]]
    ) %>% 
    filter(
        (gest_age == 25 & (wgt >= 0.6 & wgt <= 1)) | 
        (gest_age == 26 & (wgt >= 0.6 & wgt <= 1.3)) |
        (gest_age == 27 & (wgt >= 0.6 & wgt <= 1.6)) |
        (gest_age == 28 & (wgt >= 0.7 & wgt <= 1.8)) |
        (gest_age == 29 & (wgt >= 0.7 & wgt <= 1.9)) |
        (gest_age == 30 & (wgt >= 0.7 & wgt <= 2)) |
        (gest_age == 31 & (wgt >= 0.8 & wgt <= 2.3)) |
        (gest_age == 32 & (wgt >= 0.8 & wgt <= 2.4)) |
        (gest_age == 33 & (wgt >= 0.8 & wgt <= 2.6)) |
        (gest_age == 34 & (wgt >= 1.0 & wgt <= 2.9)) |
        (gest_age == 35 & (wgt >= 1.1 & wgt <= 3.2)) |
        (gest_age == 36 & (wgt >= 1.2 & wgt <= 4.0)) |
        (gest_age == 37 & (wgt >= 1.2 & wgt <= 4.0)) |
        (gest_age == 38 & (wgt >= 1.5 & wgt <= 5.0)) |
        (gest_age == 39 & (wgt >= 1.5 & wgt <= 5.0)) |
        (gest_age == 40 & (wgt >= 1.5 & wgt <= 5.0)) |
        (gest_age == 41 & (wgt >= 1.6 & wgt <= 5.0)) |
        (gest_age == 42 & (wgt >= 1.7 & wgt <= 5.0)) |
        (gest_age == 43 & (wgt >= 1.8 & wgt <= 5.0)) |
        (gest_age == 44 & (wgt >= 1.8 & wgt <= 5.0)) |
        (gest_age == 45 & (wgt >= 1.8 & wgt <= 5.0)) |
        (gest_age == 46 & (wgt >= 1.8 & wgt <= 5.0)) 
    ) %>% 
    arrange(gest_age, wgt)
```


#### Males

```{r}
tbl_1 %>% 
    pivot_wider(names_from = gest_age, values_from = survival
    ) %>% 
    gt::gt(rowname_col = "wgt") %>% 
    gt::sub_missing(columns = everything(), rows = everything(), missing_text = ""
    ) %>% 
    gt::tab_header(
        title = gt::md("**Estimated Gestational Age (weeks)**")) %>% 
     gt::tab_stubhead(gt::md("**Weight (kgs)**"))  %>% 
    gt::opt_stylize(style = 2)
```


```{r}
glm_3 <- 
    glm(outcome ~ wt*gestage, data = df_females, family = binomial)
```

```{r}
temp <- 
    glm_3 %>% 
    ggeffects::ggpredict(terms = c("wt[all]", "gestage[all]")) 
```

```{r}
tbl_1 <- 
    data.frame(
    wgt = temp[[1]],
    survival = round(temp[[2]]*100,1),
    gest_age = temp[[6]]
    ) %>% 
    filter(
        (gest_age == 25 & (wgt >= 0.6 & wgt <= 1)) | 
        (gest_age == 26 & (wgt >= 0.6 & wgt <= 1.3)) |
        (gest_age == 27 & (wgt >= 0.6 & wgt <= 1.6)) |
        (gest_age == 28 & (wgt >= 0.7 & wgt <= 1.8)) |
        (gest_age == 29 & (wgt >= 0.7 & wgt <= 1.9)) |
        (gest_age == 30 & (wgt >= 0.7 & wgt <= 2)) |
        (gest_age == 31 & (wgt >= 0.8 & wgt <= 2.3)) |
        (gest_age == 32 & (wgt >= 0.8 & wgt <= 2.4)) |
        (gest_age == 33 & (wgt >= 0.8 & wgt <= 2.6)) |
        (gest_age == 34 & (wgt >= 1.0 & wgt <= 2.9)) |
        (gest_age == 35 & (wgt >= 1.1 & wgt <= 3.2)) |
        (gest_age == 36 & (wgt >= 1.2 & wgt <= 4.0)) |
        (gest_age == 37 & (wgt >= 1.2 & wgt <= 4.0)) |
        (gest_age == 38 & (wgt >= 1.5 & wgt <= 5.0)) |
        (gest_age == 39 & (wgt >= 1.5 & wgt <= 5.0)) |
        (gest_age == 40 & (wgt >= 1.5 & wgt <= 5.0)) |
        (gest_age == 41 & (wgt >= 1.6 & wgt <= 5.0)) |
        (gest_age == 42 & (wgt >= 1.7 & wgt <= 5.0)) |
        (gest_age == 43 & (wgt >= 1.8 & wgt <= 5.0)) |
        (gest_age == 44 & (wgt >= 1.8 & wgt <= 5.0)) |
        (gest_age == 45 & (wgt >= 1.8 & wgt <= 5.0)) |
        (gest_age == 46 & (wgt >= 1.8 & wgt <= 5.0)) 
    ) %>% 
    arrange(gest_age, wgt)
```

#### Females

```{r}
tbl_1 %>% 
    pivot_wider(names_from = gest_age, values_from = survival
    ) %>% 
    gt::gt(rowname_col = "wgt") %>% 
    gt::sub_missing(columns = everything(), rows = everything(), missing_text = ""
    ) %>% 
    gt::tab_header(
        title = gt::md("**Estimated Gestational Age (weeks)**")) %>% 
     gt::tab_stubhead(gt::md("**Weight (kgs)**")) %>% 
    gt::opt_stylize(style = 2)
```

