---
title: "Newborn survival"
description: |
  Determining the survival of newborns after birth is essential for not only the pediatricians but also obstetricians. This post outlines the survival probability of newborns admitted to a Mother and Babies Unit of a tertiary hospital in Ghana.   
author:
  - name: Samuel Blay Nguah
    url: https://sbnguah.netlify.app/
date: 2023-09-22
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
```

### Background 
Predicting the survival of newborns is crucial in the clinical decision making especially during deliveries. For instance, the probability of survival of a child born at any gestational age, weight, and a specific locality should be readily known to make clinical judgments and for counselling prospective mothers.  

### Objective
This post analyses and outlines the probability of survival for newborns admitted to a tertiary hospital in Ghana. 

### Data
Data for all newborns admitted to the hospital's Mother and Baby Unit was recorded from 2013 to 2015. The probability of survival with the 95% confidence interval are as displayed below for males and female seperately.

```{r}
df_newborns <-
    read.delim("C:\\Dataset\\newborns.txt") %>% 
    select(wt, outcome, sex) %>% 
    filter(outcome %in% c("Died","Discharged"), sex != "Missing", wt > 0.9 & wt < 9) %>% 
    mutate(across(c(outcome, sex), ~factor(.)),
           wt = round(wt, 1))
```


```{r}
glm1 <- 
    df_newborns %>% 
    glm(outcome ~ wt*sex, family = "binomial", data = .)
```

```{r}
xx <- 
    glm1 %>% 
    ggeffects::ggpredict(terms = c("wt[all]", "sex"))

data.frame(
    weight = xx$x, 
    prob = round(xx$predicted*100, 1), 
    conf_low = round(xx$conf.low*100, 1), 
    conf_high = round(xx$conf.high*100,1), 
    sex = xx$group) %>% 
    pivot_wider(
        id_cols = weight, 
        names_from = sex, 
        values_from = c(prob, conf_low, conf_high)) %>%
    mutate(
        `Birth Weight (kgs)` = sprintf(fmt = "%.1f", weight),
        Female = paste0(
            sprintf(fmt = "%.1f", prob_Female), " [", 
            sprintf(fmt = "%.1f", conf_low_Female), " - ", 
            sprintf(fmt = "%.1f", conf_high_Female), "]"),
        Male = paste0(
            sprintf(fmt = "%.1f", prob_Male), " [", 
            sprintf(fmt = "%.1f", conf_low_Male), " - ", 
            sprintf(fmt = "%.1f", conf_high_Male), "]")) %>% 
    select(`Birth Weight (kgs)`, Female, Male) %>% 
    as_tibble() %>% 
    DT::datatable(
        rownames = FALSE)
    ```

This is also illustrated graphically below

```{r}
glm1 %>% 
    ggeffects::ggpredict(terms = c("wt[all]", "sex")) %>% 
    plot() +
    labs(title = "Percentage survival for inborn newborns at a teaching hospital in Ghana",
         x = "Weight in kilograms",
         y = "Probability of survival",
         color = "Sex")+
    theme_bw()
    
```





